New Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2024-10-14 jso.tex
go though the paper before submission

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{assumption}{\protect\assumptionname}
\newtheorem{claim}{\protect\claimname}
\newtheorem{condition}{\protect\conditionname}
\newtheorem{conjecture}{\protect\conjecturename}
\newtheorem{cor}{\protect\corollaryname}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{example}{\protect\examplename}
\newtheorem{lem}{\protect\lemmaname}
\newtheorem{notation}{\protect\notationname}
\newtheorem{problem}{\protect\problemname}
\newtheorem{prop}{\protect\propositionname}
\newtheorem{rem}{\protect\remarkname}
\newtheorem{thm}{\protect\theoremname}
\providecommand{\assumptionname}{Assumption}
\providecommand{\claimname}{Claim}
\providecommand{\conditionname}{Condition}
\providecommand{\conjecturename}{Conjecture}
\providecommand{\corollaryname}{Corollary}
\providecommand{\definitionname}{Definition}
\providecommand{\examplename}{Example}
\providecommand{\lemmaname}{Lemma}
\providecommand{\notationname}{Notation}
\providecommand{\problemname}{Problem}
\providecommand{\propositionname}{Proposition}
\providecommand{\remarkname}{Remark}
\providecommand{\theoremname}{Theorem}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Projection pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974),
Huber (1985)) is a dimension reduction technique aimed at identifying
informative linear projections of data. This is useful for exploring
high-dimensional data, and creating plots of the data that reveal the
main features to use for publication. The method involves optimising an
objective function known as the PP index (e.g. Hall (1989), Cook, Buja,
and Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criterion for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). The projection \(Y = XA\) is a
linear transformation that maps data from a \(p\)-dimensional space into
a \(d\)-dimensional space. The index function
\(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar function
that measures an interesting aspect of the projected data, such as
deviation from normality, presence of clusters, non-linear structure, or
other features of interest. For a fixed sample of data, PP finds the
orthonormal basis \(A\) that maximises the index value of the
projection, \(Y = XA\):

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of the PPGT, reviews existing optimisers and
index functions in the literature. Section~\ref{sec-PP-properties}
introduces the metrics that measure different properties of PP indexes,
smoothness and squintability, and Section~\ref{sec-JSO} describes the
new JSO to be used for the PPGT. Section~\ref{sec-sim-deets} summarises
the results of two simulation experiments done to assess JSO's
performance: one comparing JSO's performance improvements relative to an
existing optimiser, Creeping Random Search (CRS), and the other studying
the impact of different PP index properties on optimisation performance.
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-discussion} discusses implementation details and
insights for practitioners. Section~\ref{sec-conclusion} summarises the
work and provides suggestions for future directions.

@@

Here the formal definitions of projection distance and squint angle are
given, before the definition of squintability. Two approaches to compute
this metric numerically are described.

\begin{defn}[projection distance]\label{def:proj-dist}
Recall that $A \in \mathbb{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix. Let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance $r(A, A^*)$ between $A$ 
and $A^*$ is defined as 
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}

\begin{defn}[squint angle]\label{def:squint-angle}
Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space, aiming to balance exploration
(searching new areas) and exploitation (focusing on promising areas).
The algorithm aims to find the optimal solution by adapting the
behaviour of jellyfish to navigate towards the best solution over
iterations (Chou and Truong 2021).

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, breakable, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, titlerule=0mm, toprule=.15mm, left=2mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colback=white, leftrule=.75mm, toptitle=1mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, coltitle=black]

@@

\section{Assessing optimisers: a simulation study}\label{sec-sim-deets}

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) (Zhang et al. 2021; Laa and Cook 2020) used in the
PP guided tour to explore JSO's behaviour under different
hyper-parameter and data dimension combinations. The second simulation
studies the effect of index properties (smoothness and squintability),
along with JSO hyper-parameters, and data dimension, on the success rate
of the JSO performance. This section describes the simulation details,
with the results deferred to Section~\ref{sec-sim-res}.

@@

The performance of JSO is investigated both in comparison to the
existing optimizer, CRS, and across various hyper-parameter values. The
performance is measured by the success rate, defined as the proportion
of simulations that achieves a final index value within 0.05 of the best
index value found among all 50 simulations (see
Figure~\ref{fig-success-rate} for an illustration). This comparison is
based on projection pursuit to find the pipe shape investigated by Laa
and Cook (2020) using the \texttt{holes} index.

Fifty simulations are conducted with both JSO and CRS, in four different
data dimensions (\(d = 6, 8, 10, 12\)). JSO uses 100 jellyfish with a
maximum of 100 tries, while the CRS allows a maximum of 1000 samples at
each iteration before the algorithm terminates. The different numbers
account for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in Zhang et al. (2021) for assessing JSO, where the
design parameters are stored along with index value, projection basis,
random seed, and computation time.

Fifty additional simulations are conducted for each hyper-parameter
combination to analyze how they affect the JSO success rate. This
includes variations in the number of jellyfish (20, 50, and 100) and the
maximum number of tries (50 and 100).

@@

\subsection{Factors affecting JSO success rate: index properties and
jellyfish hyper-parameters}\label{sec-app-2}

To assess JSO's performance across various scenarios, two different data
shapes, pipe and a sine wave, are investigated in 6D and 8D spaces using
six different PP indexes: \texttt{dcor2d\_2}, \texttt{loess2d},
\texttt{MIC}, \texttt{TIC}, \texttt{spline}, and \texttt{stringy}, with
varied JSO hyper-parameters. A total of 52 combinations result,
comprising of 24 computed on the pipe data and 28 on the sine-wave data.
Again, JSO is run 50 times to calculate the success rate for each
projection pursuit.

@@

To construct a relationship among success rate, index properties
(smoothness and squintability), and jellyfish hyper-parameters, a
generalised linear model is fitted using a binomial family and a logit
link function. The data is pre-processed by 1) scaling the JSO
hyper-parameters by a factor of 10 for interpretation, 2) creating a new
binary variable \texttt{long\_time} to indicate cases with an average
run time over 20 seconds, and 3) re-coding the success rate for the
\texttt{stringy} index as 0, because none of the 50 simulations
correctly identified the sine-wave shape.

@@

\caption{\label{fig-smoothness}Illustration of steps to calculate
smoothness. For a given projection pursuit problem defined by the shape
to find, data dimension and the index function, 1) sample random bases
given the orthonormality contraint and calculate the index value
associated with each random basis, and 2) fit a Gaussian process model
of index values against projection distances to obtain the smoothness
measure.}

@@

\caption{\label{fig-squintability}Illustration of steps to calculate
squintability. For a given projection pursuit problem defined by the
shape to find, data dimension and the index function, 1) sample random
bases given the orthonormality and projection distance contraint, 2)
interpolate the sampled bases to the optimal basis and calculate the
projection distance and the index value for each interpolated basis. 3)
bin the index values by projection distances to obtain the average index
value for each bin, 4) fit the scaled sigmoid function in equation
\eqref{eq-squintability} to the binned index values against projection
distances using non-linear least square, 5) calculate the squintability
measure using equation (\ref{eq-squintability-parametric}) with
parameters estimated from the model.}

@@

Performance is assessed based on the final projections across two
optimisers (JSO and CRS) and the success rate across JSO
hyper-parameters. The final projections found by the two optimisers (JSO
and CRS) are presented in Figure~\ref{fig-proj}, broken down by 10th
quantile, faceted by the data dimensions. In the 6-dimensional data
scenario, JSO consistently identifies a clear pipe shape. The CRS also
finds the pipe shape but with a wide rim, suggesting a further polish
search may be required. With increasing dimensions, JSO may not always
identify the pipe shape due to random sampling, but it still finds the
pipe shape in over 50\% of cases. Compared to CRS, JSO achieves higher
index values and clearer pipe shapes across all quantiles in data of 8,
10, 12 dimensions, suggesting its advantage in exploring
high-dimensional spaces.

\subsection{Factors affecting JSO success rate: index properties and
jellyfish
hyper-parameters}\label{factors-affecting-jso-success-rate-index-properties-and-jellyfish-hyper-parameters}

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. The success rate calculated at each hyper-parameter
combination (number of jellyfish and the maximum number of tries) is
presented in Figure~\ref{fig-proportion}. The uncertainty is quantified
through 500 bootstrap replicates for each case. As the number of
jellyfish and maximum tries increase, the success rate also increases.
For simpler problems (6 dimensions), small parameter values (20
jellyfish and a maximum of 50 tries) can already achieve a high success
rate. However, larger parameter values (i.e.~100 jellyfish and a maximum
of 100 tries) are necessary for higher-dimensional problems (8, 10, and
12 dimensions). Increasing both parameters enhances the performance of
JSO, but it also extends the computational time required for the
optimisation, which can be computationally intensive when evaluating the
index function (such as scagnostic indexes) multiple times across
numerous iterations.

@@

The index properties, smoothness and squintability, numerically
characterise the complexity of projection pursuit index, and hence
potential optimisation problems.
Table~\ref{tbl-smoothness-squintability} presents the parameters for
calculating both metrics estimated from the Gaussian process (variance,
range, smooth, and nugget) and the scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for each case considered in the
simulation. The column \(\nu\) is used as the smoothness metric and the
column \(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

To use JSO for PPGT in the \texttt{tourr} package, specify
\texttt{search\_f\ =\ search\_jellyfish} in the guided tour. Since
multiple tour paths are generated by the jellyfish, the animation
function \texttt{animate\_*()} will not render the tour path for JSO. To
visualise path visited by each individual jellyfish, assign the
animation to an object (\texttt{res\ \textless{}-\ animate\_xy(...)}).
This will save the bases visited by the optimiser, along with relevant
metadata, as a tibble data object (see Zhang et al. (2021) for more
details on the data object). The bases visited for individual jellyfish
can then be extracted and viewed using the \texttt{planned\_tour()}
function.

\subsection{Computing the metrics for your new
index}\label{computing-the-metrics-for-your-new-index}

The \texttt{ferrn} package (Zhang et al. 2021) provides functionality
for computing the smoothness and squintability metrics. Both cases
require sampling random bases using the \texttt{sample\_bases()}
function, followed by calculating the specific metric with either
\texttt{calc\_smoothness()} or \texttt{calc\_squintability()}.

@@

To sample bases for calculating smoothness, four arguments are required:
the index function, the dataset, the number of random bases to sample,
and the best basis to compute projection distance. The output of
\texttt{sample\_bases()} is a data frame containing columns/
list-columns of sampled basis matrix, calculated projection distance and
index value. Parallellasation is available through the
\texttt{parallel\ =\ TRUE} argument to speed up the computation.

The \texttt{calc\_smoothness()} function takes the output from
\texttt{sample\_bases()} and fits a Gaussian process model to the index
values against the projection distances to compute the smoothness
metric. Starting parameters and additional Gaussian process arguments
can be specified and details of the fit can be accessed through the
\texttt{fit\_res} attribute of the \texttt{calc\_smoothness()} output.

@@

Sampling bases for calculating squintability includes an additional step
of interpolating between the sampled bases and the best basis. This is
specified through setting the \texttt{step\_size} and
\texttt{min\_proj\_dist} arguments in the \texttt{sample\_bases()}
function to non-NA numerical values. Since the projection distance
typically ranges from 0 to 2, it is recommended to set
\texttt{step\_size} to 0.1 or lower, and \texttt{min\_proj\_dist} to be
at least 0.5 to ensure a meaningful interpolation length.

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit. The output provides the
estimated parameters for the scaled sigmoid function (\(\theta_1\) to
\(\theta_4\)) and the squintability metric for the parametric case. For
the non-parametric case, it shows the maximum gradient attained
(\texttt{max\_d}), the corresponding projection distance
(\texttt{max\_dist}), and the squintability metric as their products.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{prop}{\protect\propositionname}
\providecommand{\definitionname}{Definition}
\providecommand{\propositionname}{Proposition}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Projection pursuit (PP) (Kruskal (1969), Friedman and Tukey (1974),
Huber (1985)) is a dimension reduction technique aimed at identifying
informative linear projections of data. This is useful for exploring
high-dimensional data, and creating plots of the data that reveal the
main features to use for publication. The method involves optimising an
objective function known as the PP index (e.g. Hall (1989), Cook, Buja,
and Cabrera (1993), Lee and Cook (2010), Loperfido (2018), Loperfido
(2020)), which defines the criterion for what constitutes interesting or
informative projections. Let \(X \in \mathbb{R}^{n\times p}\) be the
data matrix, \(A \in\mathbb{R}^{p \times d}\) be an orthonormal matrix,
where \(A\) belongs to the Stiefel manifold
\(\mathcal{A} = V_d(\mathbb{R}^p)\). The projection \(Y = XA\) is a
linear transformation that maps data from a \(p\)-dimensional space into
a \(d\)-dimensional space. The index function
\(f(XA): \mathbb{R}^{n \times d} \to \mathbb{R}\) is a scalar function
that measures an interesting aspect of the projected data, such as
deviation from normality, presence of clusters, or non-linear structure.
For a fixed sample of data, PP finds the orthonormal basis \(A\) that
maximises the index value of the projection, \(Y = XA\):

@@

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of the PPGT, reviews existing optimisers and
index functions in the literature. Section~\ref{sec-PP-properties}
introduces the metrics that measure two properties of PP indexes,
smoothness and squintability, and Section~\ref{sec-JSO} describes the
new JSO to be used for the PPGT. Section~\ref{sec-sim-deets} outlines
two simulation experiments to assess JSO's performance: one comparing
JSO's performance improvements relative to an existing optimiser,
Creeping Random Search (CRS), and the other studying the impact of PP
index properties on optimisation performance, and
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-discussion} discusses the implementation of JSO in the
\texttt{tourr} package and the PP property calculation in the
\texttt{ferrn} package. Section~\ref{sec-conclusion} summarises the work
and provides suggestions for future directions.

@@

This section defines the projection distance and introduces the
squintability metric, followed by a description of two numerical
approaches for computing squintability.

\begin{defn}[projection distance]\label{def:proj-dist}
Let $A \in \mathbb{R}^{p \times d}$ is a $d$-dimensional orthonormal matrix, and let $A^*$ be the optimal matrix that achieves the maximum index value for a given data. The projection distance between $A$ 
and $A^*$, $r(A, A^*)$, is defined by
$r(A, A^*) = \lVert AA^\prime - A^*A^{*\prime}\,\rVert _F$
where $\lVert . \rVert _F$ denotes the Frobenius norm, given by
$\lVert M \rVert _F = \sqrt{\sum_{ij} M_{ij}^2}$. 
\end{defn}
\begin{defn}[squint angle]\label{def:squint-angle}

Let $A$ and $B$ be two $d$-dimensional orthonormal matrices in $\mathbb{R}^p$. The squint angle $\theta$ between the subspace spanned by $A$ and $B$ is defined as the smallest principal angle between these subspaces: $\theta = \min_{i \in \{1, \cdots, d\}} \arccos(\tau_i)$, where $\tau_i$ are the singular values of the matrix $M = A^T B$ obtained from its singular value decomposition.
\end{defn}

@@

The Jellyfish Search Optimiser (JSO) mimics the natural movements of
jellyfish, which include passive and active motions driven by ocean
currents and their swimming patterns, respectively. In the context of
optimization, these movements are abstracted to explore the search
space, aiming to balance exploration (searching new areas) and
exploitation (focusing on promising areas). The algorithm aims to find
the optimal solution by adapting the behaviour of jellyfish to navigate
towards the best solution over iterations (Chou and Truong 2021).

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, opacityback=0, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, toprule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6, rightrule=.15mm]

@@

\section{Assessing optimisers: two simulation
studies}\label{sec-sim-deets}

This section describes the details of two simulation studies: one
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) (Zhang et al. 2021; Laa and Cook 2020) to explore JSO's
behaviour under different data dimensions. The second simulation first
examines the performance JSO under different JSO hyper-parameters
(number of jellyfish and the maximum number of tries), and then studies
the effect of index properties (smoothness and squintability), along
with JSO hyper-parameters, and data dimension, on the performance of JSO
through a logistic regression.

The performance of JSO is measured by the success rate, defined as the
proportion of simulations that achieves a final index value within 0.05
of the best index value found among all 50 simulations (see
Figure~\ref{fig-success-rate} for an illustration). This comparison is
based on projection pursuit to find the pipe shape investigated by Laa
and Cook (2020) using the \texttt{holes} index.

@@

The JSO performance is investigated both in comparison to the existing
optimizer, CRS, in a pipe-finding problem using the holes index in the
6, 8, 10, and 12-dimensional space. Fifty simulations are conducted in
four data dimensions (\(d = 6, 8, 10, 12\)) with each optimiser. JSO
uses 100 jellyfish with a maximum of 100 tries, while the CRS allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers account for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in Zhang et al. (2021) for
assessing JSO, where the design parameters are stored along with index
value, projection basis, random seed, and computation time.

@@

\subsection{Factors affecting JSO success rate: JSO hyper-parameters and
index properties}\label{sec-app-2}

To examine JSO's performance across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. The success rate is
calculated for each combination.

To assess JSO's performance across various PP problems, two different
data shapes, pipe and a sine wave, are investigated in 6D and 8D spaces
using six different PP indexes: \texttt{dcor2d\_2}, \texttt{loess2d},
\texttt{MIC}, \texttt{TIC}, \texttt{spline}, and \texttt{stringy}, under
varied JSO hyper-parameters. This results in a total of 52 scenarios,
comprising of 24 computed on the pipe data and 28 on the sine-wave data.
Again, JSO is run 50 times for each scenario to calculate the success
rate for each PP problem.

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 20
seconds, and 3) re-coding the success rate for the \texttt{stringy}
index as 0, because none of the 50 simulations correctly identified the
sine-wave shape.

@@

\caption{\label{fig-smoothness}Steps for calculating smoothness in a
projection pursuit problem. Given the target shape, data dimension and
the index function: (1) sample random bases given the orthonormality
contraint and calculate their corresponding index values, and (2) fit a
Gaussian process model of index values against the sampled bases to
obtain the smoothness measure.}

@@

\caption{\label{fig-squintability}Steps for calculating squintability in
a projection pursuit problem. Given the target shape, data dimension and
the index function: (1) sample random bases given the orthonormality and
projection distance contraint, (2) interpolate the sampled bases to the
optimal basis to calculate the projection distance and the index value.
(3) bin the index values by projection distances to obtain the average
index value for each bin, (4) fit the scaled sigmoid function in
equation \eqref{eq-squintability} to the binned index values against
projection distances using non-linear least square, 5) calculate the
squintability measure using equation (\ref{eq-squintability-parametric})
with parameters estimated from the model.}

@@

Performance of JSO is assessed based on the final projections across the
two optimisers: JSO and CRS. The final projections found by the two
optimisers are presented in Figure~\ref{fig-proj}, broken down by 10th
quantile, faceted by the data dimensions. In the 6-dimensional data
scenario, JSO consistently identifies a clear pipe shape. The CRS also
finds the pipe shape but with a wide rim, suggesting a further polish
search may be required. With increasing dimensions, JSO may not always
identify the pipe shape due to random sampling, but it still finds the
pipe shape in over 50\% of cases. Compared to CRS, JSO achieves higher
index values and clearer pipe shapes across all quantiles in data of 8,
10, 12 dimensions, suggesting its advantage in exploring
high-dimensional spaces.

\subsection{Factors affecting JSO success rate: JSO hyper-parameters and
index
properties}\label{factors-affecting-jso-success-rate-jso-hyper-parameters-and-index-properties}

The effect of JSO hyper--parameters (number of jellyfish and the maximum
number of tries) on the success rate is presented in
Figure~\ref{fig-proportion}. The uncertainty is quantified through 500
bootstrap replicates for each case. As the number of jellyfish and
maximum tries increase, the success rate also increases. For simpler
problems (6 dimensions), small parameter values (20 jellyfish and a
maximum of 50 tries) can already achieve a high success rate. However,
larger parameter values (i.e.~100 jellyfish and a maximum of 100 tries)
are necessary for higher-dimensional problems (8, 10, and 12
dimensions). Increasing both parameters enhances the performance of JSO,
but it also extends the computational time required for the
optimisation, which can be computationally intensive when evaluating the
index function (such as scagnostic indexes) multiple times across
numerous iterations.

Smoothness and squintability are calculated across a collection of
pipe-finding and sine-wave finding problems to construct the
relationship between success rate, JSO hyper-parameters, and index
properties. Table~\ref{tbl-smoothness-squintability} presents the
parameters estimated from the Gaussian process (outputscale \(\eta\),
lengthscale \(\ell\), smoothness \(\nu\), and nugget \(\sigma\)) and
from the scaled logistic function (\(\theta_1\) to \(\theta_4\)) for
calculating smoothness and squintability in each PP problem considered.
The column \(\nu\) is used as the smoothness metric and the column
\(\varsigma\) is calculated as equation
(\ref{eq-squintability-parametric}) as the squintability metric.

@@

@@

To use JSO for PPGT in the \texttt{tourr} package, specify
\texttt{search\_f\ =\ search\_jellyfish} in the guided tour. Unlike
existing optimisers, the animation function \texttt{animate\_*()} won't
directly render the tour path for JSO due to the generation of multiple
tour paths. To visualise the path visited by each individual jellyfish,
assign the animation to an object
(\texttt{res\ \textless{}-\ animate\_xy(...)}). This will save the bases
visited by JSO, along with relevant metadata, as a tibble data object
(see Zhang et al. (2021) for more details on the data object). The bases
visited for individual jellyfish can then be extracted and viewed using
the \texttt{planned\_tour()} function.

\subsection{Computing the index properties for your new
index}\label{computing-the-index-properties-for-your-new-index}

The \texttt{ferrn} package (Zhang et al. 2021) provides functionality
for computing the smoothness and squintability metrics. Both metrics
require sampling random bases using the \texttt{sample\_bases()}
function, followed by the metric calculation with
\texttt{calc\_smoothness()} or \texttt{calc\_squintability()}.

@@

To sample bases for calculating smoothness, the following arguments are
required: the index function, the dataset, and the number of random
bases to sample. The output of \texttt{sample\_bases()} is a data frame
with a list-column of sampled basis matrix and index value.
Parallellasation is available to speed up the index value computation
through the \texttt{parallel\ =\ TRUE} argument.

The \texttt{calc\_smoothness()} function takes the output from
\texttt{sample\_bases()} and fits a Gaussian process model to the index
values against the sampled basis, as the location, to obtain the
smoothness metric. Starting parameters and additional Gaussian process
arguments can be specified and details of the fit can be accessed
through the \texttt{fit\_res} attribute of the output.

@@

Bases sampling for squintability includes an additional step of
interpolating between the sampled bases and the best basis. This step is
performed when the arguments \texttt{step\_size} and
\texttt{min\_proj\_dist} in the \texttt{sample\_bases()} function are
set to non-NA numerical values. Given the projection distance typically
ranging from 0 to 2, it is recommended to set \texttt{step\_size} to 0.1
or lower, and \texttt{min\_proj\_dist} to be at least 0.5 to ensure a
meaningful interpolation length.

The \texttt{calc\_squintability()} function computes squintability using
two methods: 1) parametrically, by fitting a scaled sigmoid function
through non-linear least square (\texttt{method\ =\ "nls"}), and 2)
non-parametrically, using kernel smoothing (\texttt{method\ =\ "ks"}). A
\texttt{bin\_width} argument is required to average the index values
over the projection distance before the fit. For the parametric case,
the output provides the estimated parameters for the scaled sigmoid
function (\(\theta_1\) to \(\theta_4\)) and the calculated squintability
metric as Equation \ref{eq-squintability-parametric}. For the
non-parametric case, it shows the maximum gradient attained
(\texttt{max\_d}), the corresponding projection distance
(\texttt{max\_dist}), and the squintability metric as their products.
